Bayesian Learning of Degenerate Linear~Gaussian~State~Space~Models using~Markov~Chain~Monte~Carlo

### 2016-01-14 notes/reviews/response_to_reviews.tex

\begin{review}
It will be helpful to provide some relevant applications in Section I, explaining why a degenerate model will be more appropriate.
\end{review}

\begin{response}
We have modified the introduction to mention the example of tracking multiple points on a rigid body.
\end{response}

\begin{review}
There is no complexity analysis.
\end{review}

\begin{response}
There is a brief discussion of the computational complexity in the final section of the paper. There is not much to say -- it is of the same order for the full rank and degenerate models but with a higher cost per iteration. We do not see much value in including a detailed operation count, since the dominating factor when considering processing time will be the difference in autocorrelation and the resulting difference in the necessary length of the chain, which will vary by model and data set.
\end{response}

@@

\section*{Response to Reviewer 3}

\begin{review}
In this paper, the authors address the Bayesian learning of degenerate linear and Gaussian state space models. In this specific context, classical Gibbs sampler can not be used and therefore the authors propose an interesting and complete framework based on Markov Chain Monte Carlo (MCMC) to deal with this challenging problem. The paper is well written and organized. I have the following remarks:

Even if I found the paper well organized, I think that the authors should summarize the complete framework (RJ-MCMC) in an algorithm - in my opinion it will help the reader to understand it by summarizing all the steps of the proposed scheme in one place.
\end{review}

\begin{response}
 A good point. We have added algorithm pseudo-code describing the steps of the complete MCMC kernel. It was an oversight to leave this out.
\end{response}

@@

\begin{review}
In Section III-C, the authors propose to use a collapsed Gibbs move to more efficiently exploit the target on the space of $F$ and $Q$, alternatively. I think the authors should discuss in more details why such collapsed Gibbs is important in the text. Moreover, the importance of such steps could be easily illustrated in the results section so the reader can appreciate the benefit of using such a strategy.
\end{review}

\begin{response}
 There is a misunderstanding here, probably due to the confusing terminology. The ``collapsed Gibbs move'' simply means the composition of first targeting $\pi(F, Q)$ with Metropolis-Hastings followed by a draw from $\pi(x_{1:T}|F, Q)$. These moves are essential for the sampler to be irreducible, and they are woefully inefficient. To speed things up, we also include exact draws from the the parameter conditional $\pi(F, Q|x_{1:T})$. These are inspired by the Gibbs sampler for the full rank algorithm, but cannot fully explore the parameter space, so we refer to them as constrained Gibbs moves.

 It does not make sense to run the sampler without the Metropolis-Hastings/constrained-Gibbs moves. It will be stuck in whatever subspace we happen to initialise it in. We {\em could} run it without the draws from the parameter conditional, but our python implementation would take literally years to converge.

 We have substantially re-arranged section III and added further explanation of these concepts. We hope it is clearer now.
\end{response}

@@

\begin{review}
There are no figures that show the evolution of the Markov chain OR the posterior approximation of the unknown rank $r$
The authors wrote that ``the sampler identifies the true rank.... and remains at this value thereafter'' but this can be worrying - this statement could also mean that the sampler is not mixing properly for this variable $r$. The author should be more precised by saying ``Trace plots suggest that the sampler converges within a few thousand iterations. See figure 2'' - The convergence showed in this figure is only related to the variable $\xi_y$.
\end{review}

\begin{response}
 \meta{Explain this and improve in the paper. Add simple example of changing rank to show that it is not stuck.}
\end{response}

\begin{review}
The authors could illustrate in the first toy example the benefit of the collapsed Gibbs
moves by showing the performance of the sampler with vs without this strategy.
\end{review}

\begin{response}
 \meta{Try running without parameter conditional draws on toy model. Very slow, right?!}
\end{response}

@@

\begin{review}
I personally think that the proposed algorithm should be illustrated with at least a different set of parameters for the toy example (e.g. different values for $\xi_y$ thus allowing to have a curve of RMSE vs $\xi_y$ )
\end{review}

\begin{response}
 \meta{Ok... if it will make you happy. Obviously it will work for some parameters and not others.}
\end{response}

\begin{review}
List of minor remarks/typos:

Just after Eq. (16), the authors should be more precise on the following sentence to avoid any confusion: ``However, whereas before $Q$ was required to be positive definite, this is no longer the case'' (I would add) for unknown $G$.
\end{review}

\begin{response}
 We have re-written this sentence. 
\end{response}

\begin{review}
Please correct in Section V first sentence ... known a prior. by ... known a priori.
\end{review}

\begin{response}
This has been corrected.
\end{response}

### 2016-01-14 notes/paper-degenerate-linear-systems.tex
More corrections: Add pseudo-code. Re-arranged degenerate model section.

\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algorithmic}
% \usepackage{graphicx}
\usepackage{hyperref}
\usepackage[caption=false]{subfig}
\usepackage{cite}

@@

@@

Our principal interest is to learn the system matrices $\lgtm$ and $\lgtv$ from a sequence of observations $\ob{1:\timax}$. The posterior distribution over these variables $\den(\lgtm, \lgtv | \ob{1:\timax})$ is not amenable to MCMC sampling because there is no analytic expression for the density. Instead we introduce the latent state sequence $\ls{1:\timax}$ as a nuisance variable and target the joint posterior,
%
\begin{IEEEeqnarray}{rCl}
 \postden(\lgtm, \lgtv, \ls{1:\timax}) & = & \den(\lgtm, \lgtv, \ls{1:\timax} | \ob{1:\timax}) \\
 & \propto & \den(\ob{1:\timax}|\ls{1:\timax}) \den(\ls{1:\timax}|\lgtm,\lgtv) \den(\lgtm|\lgtv) \den(\lgtv) \nonumber      .
\end{IEEEeqnarray}

@@

A commonly encountered variation on the basic linear Gaussian state space model is as follows,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \lgdm \tn{\ti} \label{eq:degenerate_transition} \\
 \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\idmat}      ,
\end{IEEEeqnarray}
%
in which $\tn{\ti}$ as dimension $\rk$, and $\lgdm$ is $\lsd\times\rk$. This is almost equivalent to the previous case, with $\lgtv=\lgdm\lgdm\tr$. However, whereas before $\lgtv$ was required to be positive definite, this is no longer the case. When this occurs, the conventional transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgdm)$ is not defined, and the sampling operations underlying our Gibbs sampler are no longer valid. Degenerate models such as these arise in tracking applications (see \cite{Maskell2004} for a discussion of this phenomenon); when non-Markovian models, such as autoregressive process, are converted to state space form; and when it is natural to parameterise a system with more latent states than there are degrees of freedom (see section~\ref{sec:mocap}). In this section, we introduce suitable parameterisations for such degenerate transition models, and show how MCMC moves may be implemented for learning. Here we treat $\rk$ as fixed and known; in the following section mechanisms will be introduced to learn this too.

@@

There is an obvious way to avoid being constrained to a particular subspace by the sampled state trajectory; simply return to targeting the marginal posterior $\postden(\lgtm, \lgtv|\ob{1:\timax})$ and fall back to using a generic Metropolis-Hastings approach. However, with a complex and high dimensional parameter space, designing efficient proposal distributions for Metropolis-Hastings is difficult. We would prefer to attain the automated, fast-mixing properties displayed by the Gibbs sampler on the full rank model. We will show how a Gibbs kernel may still be applied to sample a component of the parameters, with the remainder handled using Metropolis-Hastings.

@@

Note that the forward-filtering-backward-sampling procedure can be used as before to draw from the state conditional. See appendix~\ref{app:state-sampling}. The parameter conditional, on the other hand, will require some modifications.

The likelihood function for the entire state trajectory is thus,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm, \lgtv) \propto \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \nonumber \\
 &\propto& \determ{\tvfull}^{-\half(\timax-1)} \exp\Bigg( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \nonumber \\
 && \qquad\qquad\qquad\qquad\qquad\qquad\qquad \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \Bigg) \nonumber \\
 %&=& \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
 %&=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
 &=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\bigg(-\half\trace \big[ \tvfull\inv ( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad -\: \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot ) \big] \bigg)
\end{IEEEeqnarray}
%
in which the sufficient statistics are as before and,
%
\begin{equation}
 \lgtmrot = \tvrot\tr \lgtm      .
\end{equation}

Now introduce,
%
\begin{IEEEeqnarray}{rCl}
 \lgtmorth &=& (\idmat-\tvrot\tvrot\tr) \lgtm     ,
\end{IEEEeqnarray}
%
such that the transition matrix may be decomposed into orthogonal components,
%
\begin{IEEEeqnarray}{rCl}
 \lgtm &=& \tvrot \lgtmrot + \lgtmorth    .
\end{IEEEeqnarray}
%
Using this, the constraint equation becomes,
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& (\tvrot \lgtmrot + \lgtmorth) \LS{1:\timax-1} + \tvrot Z \nonumber \\
 \Rightarrow \LS{2:\timax} &=& \lgtmorth \LS{1:\timax-1} + \tvrot Z' \nonumber      .
\end{IEEEeqnarray}
%
Notice that $\lgtmrot$ and $\tvfull$ do not appear in this equation, and thus may be freely altered by the sampler. In contrast, $\lgtmorth$ and $\tvrot$ are uniquely determined. Sampling from $\postden(\lgtm,\lgtv|\ls{1:\timax})$ thus entails drawing new values from $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$.

\subsubsection{Sampling the Conditional}
Using the constrained likelihood function and the transformed prior, we now obtain the required posterior conditional distribution $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$. This follows the same algebraic steps as in the full rank case and leads to a similar matrix normal--inverse Wishart distribution,
%
\begin{align}
 \tvfull | \tvrot, \ls{1:\timax} &\sim \iwishartdist{\postdof}{\postscalematrix} \\
 \lgtmrot | \tvfull, \tvrot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}     ,
\end{align}
%
with the following updated hyperparameters,
%
\begin{IEEEeqnarray}{rCl}
 \postcolumnvariance\inv                 &=& \priorcolumnvariance\inv + \suffstats{1} \\
 \postmeanmatrix \postcolumnvariance\inv &=& \tvrot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
 \postdof                                &=& \rk + \suffstats{0} \\
 \postscalematrix                        &=& (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\left( \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tvrot \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad -\: \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
\end{IEEEeqnarray}

%TODO: move this
\subsubsection{Forward Filtering Backward Sampling}

The forward-filtering-backward-sampling procedure for drawing from $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ can be applied to a degenerate model without modification. The Kalman filter stage is

\subsection{Metropolis-Hastings on the Posterior Marginal}

Gibbs sampling is very efficient and requires no algorithm tuning but it only allows the Markov chain to explore a subset of the parameter space. In this section we introduce two Metropolis-Hastings kernels which target the marginal posterior $\postden(\lgtm,\lgtv)$ and allow the sampler to fully explore the target distribution. These sampling operations are each followed by a draw of the state trajectory from $\postden(\ls{1:\timax}|\lgtm,\lgtv)$, thus constituting a collapsed Gibbs move \cite{Dyk2008}.

\subsubsection{Covariance Random Walk}

@@

\subsubsection{Transition Matrix Random Walk}

@@

@@

\subsection{Reversible Jump for Rank Learning}

@@

The reversible transformation can be accomplished in terms of eigenvectors and eigenvalues. Suppose the current value of the transition covariance is $\lgtv = \tvvec \tvval \tvvec\tr$, which has rank $\rk$. To increase the rank, first we sample an additional eigenvalue $\eval{\rk+1} \in [0,\eval{\rk}]$ (recall that the elements of $\tvval$ are sorted in decreasing order of magnitude, so $\eval{\rk+1} < \eval{\rk} < \dots < \eval{1}$) and a corresponding eigenvector $\evec{\rk+1} \in \evecset{\rk+1}$, from the set of vectors which could legitimately be appended to $\tvvec$ (i.e. orthonormal to the existing eigenvectors with positive sign of the first element),
%
\begin{IEEEeqnarray}{rCl}
 \evecset{\rk+1} &=& \{\evec{} : |\evec{}|=1,\:\tvvec\tr\evec{}=0,\:\sign(\evec{\rk+1,1}=+1 \}    .
\end{IEEEeqnarray}
%
A new covariance matrix is then generated using the transformation,
%
\begin{IEEEeqnarray}{rCl}
 \lgtv\mcnew &=& \lgtv + \eval{\rk+1} \evec{\rk+1}\evec{\rk+1}\tr \\
 &=& \begin{bmatrix} \tvvec & \evec{\rk+1} \end{bmatrix} \begin{bmatrix} \tvval & \zmat \\ \zmat & \eval{\rk+1} \end{bmatrix} \begin{bmatrix} \tvvec\tr \\ \evec{\rk+1}\tr \end{bmatrix}
\end{IEEEeqnarray}
%
The inverse transformation, from $\lgtv\mcnew$ to $\{\lgtv, \eval{\rk+1}, \evec{\rk+1}\}$ simply requires identifying and removing the smallest eigenvalue and its corresponding eigenvector. This is an effective choice because the component with the smallest eigenvalue has the least effect on the likelihood of the observed data, and will thus have the highest probability of being accepted.

@@

\meta{Pseudo-code here}

@@

In the previous sections, we have assumed that $\lgom$ and $\lgov$ are fixed and known a prior. In practice, $\lgov$ will often be unknown. In both the following examples we use models which have,
\begin{IEEEeqnarray}{rCl}
 \lgov &=& \xi_{\ob{}} \idmat     ,
\end{IEEEeqnarray}
%
in which $\xi_{\ob{}}$ is unknown. This additional parameter may be learned in combination with the transition model by the MCMC sampler. This entails changing the target distribution to $\postden(\lgtm, \lgtv, \xi_{\ob{}}, \ls{1:\timax})$. An additional Gibbs move is introduced to draw from $\postden(\xi_{\ob{}} | \lgtm, \lgtv, \ls{1:\timax})$. This may be done exactly if we use a conjugate inverse gamma prior.

@@

The forward-filtering-backward-sampling procedure draws a sample from $\den(\ls{1:\timax}|\ob{1:\timax}, \lgtm, \lgtv)$. This is achieved by first sampling from the final filtering distribution $\den(\ls{\timax}|\ob{1:\timax})$ and then for $\ti=\timax-1 , \dots , 1$ sampling from $\den(\ls{\ti}|\ls{\ti+1}, \ob{1:\ti}, \lgtm, \lgtv)$. Each of these factors is also a Gaussian,
%
\begin{IEEEeqnarray}{l}
 \den(\ls{\ti}|\ls{\ti+1}, \ob{1:\ti}, \lgtm, \lgtv) \nonumber \\
 \quad = \normalden{\ls{\ti}}{ \lsfmn{\ti} + K_t (\ls{\ti+1}-\lgtm\lsfmn{\ti}) }{ \lspvr{\ti} - K_t \lgtm \lsfvr{\ti} }      ,
\end{IEEEeqnarray}
%
where $K_t = \lsfvr{\ti} \lgtm\tr (\lgtm \lsfvr{\ti} \lgtm\tr + \lgtv)\inv$. The proof is straightforward using standard identities for the Gaussian distribution \cite{Sarkka2013}.

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algorithmic}
% \usepackage{graphicx}
\usepackage{hyperref}
\usepackage[caption=false]{subfig}
\usepackage{cite}
\usepackage{setspace}

@@

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

@@

Our principal interest is to learn the system matrices $\lgtm$ and $\lgtv$ from a sequence of observations $\ob{1:\timax}$. The posterior distribution over these variables $\postden(\lgtm, \lgtv) = \den(\lgtm, \lgtv | \ob{1:\timax})$ is not amenable to MCMC sampling because there is no analytic expression for the density. Instead we introduce the latent state sequence $\ls{1:\timax}$ as a nuisance variable and target the joint posterior,
%
\begin{IEEEeqnarray}{rCl}
 \postden(\lgtm, \lgtv, \ls{1:\timax}) & = & \den(\lgtm, \lgtv, \ls{1:\timax} | \ob{1:\timax}) \\
 & \propto & \den(\ob{1:\timax}|\ls{1:\timax}) \den(\ls{1:\timax}|\lgtm,\lgtv) \den(\lgtm|\lgtv) \den(\lgtv) \nonumber      .
\end{IEEEeqnarray}

@@

A commonly encountered variation on the basic linear Gaussian state space model is as follows,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \lgdm \tn{\ti} \label{eq:degenerate_transition} \\
 \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\idmat}      ,
\end{IEEEeqnarray}
%
in which $\tn{\ti}$ as dimension $\rk$, and $\lgdm$ is $\lsd\times\rk$. This is almost equivalent to the model in the previous section, with $\lgtv=\lgdm\lgdm\tr$. However, in the previous model $\lgtv$ was positive definite by definition, whereas here it may not be full rank. When this occurs, the conventional transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgdm)$ is not defined, and the sampling operations underlying our Gibbs sampler are no longer valid. Degenerate models such as these arise in tracking applications (see \cite{Maskell2004} for a discussion of this phenomenon); when non-Markovian models, such as autoregressive process, are converted to state space form; and when it is natural to parameterise a system with more latent states than there are degrees of freedom (see section~\ref{sec:mocap}). In this section, we introduce suitable parameterisations for such degenerate transition models, and show how MCMC moves may be implemented for learning. Here we treat $\rk$ as fixed and known; in the following section mechanisms will be introduced to learn this too.

@@

There is an obvious way to avoid being constrained to a particular subspace by the sampled state trajectory; simply return to targeting the marginal posterior $\postden(\lgtm, \lgtv)$ and fall back to using any generic Metropolis-Hastings approach. However, with a complex and high dimensional parameter space, designing efficient proposal distributions for Metropolis-Hastings is difficult. We would prefer to attain the automated, fast-mixing properties displayed by the Gibbs sampler on the full rank model.

The solution we propose here is to use Metropolis-Hastings steps targeting the marginal to ensure irreducibility of the sampler, combined with constrained Gibbs sampling steps, which can only modify a component of the parameters, to accelerate the mixing of the chain. As in the full rank case, the target distribution will be the posterior over parameters and the state trajectory $\postden(\lgtm, \lgtv, \ls{1:\timax})$. This means that each of the Metropolis-Hastings steps targeting the marginal posterior $\postden(\lgtm, \lgtv)$ must be followed by a draw from the state conditional $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ in order to complete the iteration. The two two steps together comprise a collapsed Gibbs move \cite{Dyk2008}. Note that the forward-filtering-backward-sampling procedure can be used without modification to draw from the state conditional. See appendix~\ref{app:state-sampling}.

\subsection{Covariance Decompositions}

@@

\subsection{Metropolis-Hastings for the Posterior Marginal}

In this section we introduce two Metropolis-Hastings kernels which target the marginal posterior $\postden(\lgtm,\lgtv)$. Sampling from each of these should be followed by a draw of the state trajectory from $\postden(\ls{1:\timax}|\lgtm,\lgtv)$. Together, this constitutes a collapsed Gibbs move for the target joint posterior distribution \cite{Dyk2008}.

These moves are needed in order to ensure that the sampler can attain any possible values in the parameter space. However, the resulting Markov chain are liable to mix very poorly, since the proposals are dumb random walks. In the following section, we will describe how to draw from the parameter conditional distribution exactly, which greatly accelerates the mixing of the chain.

\subsubsection{Covariance Random Walk} \label{sec:mh-rw-transition-covariance}

@@

\subsubsection{Transition Matrix Random Walk} \label{sec:mh-rw-transition-matrix}

@@

The likelihood function for the entire state trajectory is thus,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm, \lgtv) \propto \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \nonumber \\
 &\propto& \determ{\tvfull}^{-\half(\timax-1)} \exp\Bigg( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \nonumber \\
 && \qquad\qquad\qquad\qquad\qquad\qquad\qquad \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \Bigg) \nonumber \\
 %&=& \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
 %&=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
 &=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\bigg(-\half\trace \big[ \tvfull\inv ( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad -\: \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot ) \big] \bigg)
\end{IEEEeqnarray}
%
in which the sufficient statistics are as before and,
%
\begin{equation}
 \lgtmrot = \tvrot\tr \lgtm \label{eq:transition_matrix_free}      .
\end{equation}

Now introduce,
%
\begin{IEEEeqnarray}{rCl}
 \lgtmorth &=& (\idmat-\tvrot\tvrot\tr) \lgtm \label{eq:transition_matrix_fixed}    ,
\end{IEEEeqnarray}
%
such that the transition matrix may be decomposed into orthogonal components,
%
\begin{IEEEeqnarray}{rCl}
 \lgtm &=& \tvrot \lgtmrot + \lgtmorth \label{eq:transition_matrix_decomposition}    .
\end{IEEEeqnarray}
%
Using this, the constraint equation becomes,
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& (\tvrot \lgtmrot + \lgtmorth) \LS{1:\timax-1} + \tvrot Z \nonumber \\
 \Rightarrow \LS{2:\timax} &=& \lgtmorth \LS{1:\timax-1} + \tvrot Z' \nonumber      .
\end{IEEEeqnarray}
%
Notice that $\lgtmrot$ and $\tvfull$ do not appear in this equation, and thus may be freely altered by the sampler. In contrast, $\lgtmorth$ and $\tvrot$ are uniquely determined. Sampling from $\postden(\lgtm,\lgtv|\ls{1:\timax})$ thus entails drawing new values from $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$.

\subsubsection{Sampling the Conditional}
Using the constrained likelihood function and the transformed prior, we now obtain the required posterior conditional distribution $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$. This follows the same algebraic steps as in the full rank case and leads to a similar matrix normal--inverse Wishart distribution,
%
\begin{align}
 \tvfull | \tvrot, \ls{1:\timax} &\sim \iwishartdist{\postdof}{\postscalematrix} \nonumber \\
 \lgtmrot | \tvfull, \tvrot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance} \label{eq:constrained_parameter_conditional}      ,
\end{align}
%
with the following updated hyperparameters,
%
\begin{IEEEeqnarray}{rCl}
 \postcolumnvariance\inv                 &=& \priorcolumnvariance\inv + \suffstats{1} \\
 \postmeanmatrix \postcolumnvariance\inv &=& \tvrot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
 \postdof                                &=& \rk + \suffstats{0} \\
 \postscalematrix                        &=& (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\left( \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tvrot \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad -\: \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
\end{IEEEeqnarray}

@@

\subsection{Reversible Jump for Rank Learning} \label{sec:rjmcmc-covariance-rank}

@@

The reversible transformation can be accomplished in terms of eigenvectors and eigenvalues. Suppose the current value of the transition covariance is $\lgtv = \tvvec \tvval \tvvec\tr$, which has rank $\rk$. To increase the rank, first we sample an additional eigenvalue $\eval{\rk+1} \in [0,\eval{\rk}]$ (recall that the elements of $\tvval$ are sorted in decreasing order of magnitude, so $\eval{\rk+1} < \eval{\rk} < \dots < \eval{1}$) and a corresponding eigenvector $\evec{\rk+1} \in \evecset{\rk+1}$, from the set of vectors which could legitimately be appended to $\tvvec$ (i.e. orthonormal to the existing eigenvectors with positive sign of the first element),
%
\begin{IEEEeqnarray}{rCl}
 \evecset{\rk+1} &=& \{\evec{} : |\evec{}|=1,\:\tvvec\tr\evec{}=0,\:\sign(\evec{\rk+1,1})=+1 \}    .
\end{IEEEeqnarray}
%
A new covariance matrix is then generated using the transformation,
%
\begin{IEEEeqnarray}{rCl}
 \lgtv\mcnew &=& \lgtv + \eval{\rk+1} \evec{\rk+1}\evec{\rk+1}\tr \\
 &=& \begin{bmatrix} \tvvec & \evec{\rk+1} \end{bmatrix} \begin{bmatrix} \tvval & \zmat \\ \zmat & \eval{\rk+1} \end{bmatrix} \begin{bmatrix} \tvvec\tr \\ \evec{\rk+1}\tr \end{bmatrix}
\end{IEEEeqnarray}
%
The inverse transformation, from $\lgtv\mcnew$ to $\{\lgtv, \eval{\rk+1}, \evec{\rk+1}\}$ simply requires identifying and removing the smallest eigenvalue and its corresponding eigenvector. This is an effective choice because the component with the smallest eigenvalue has the least effect on the likelihood of the observed data, and will thus have the highest probability of being accepted.

@@

There are many ways in which the various Markov kernels could be combined to produce a complete MCMC sampler. Our suggested scheme is shown in algorithm~\ref{alg:complete-degenerate-model-algorithm}. A Markov chain is generated by repeatedly sampling according to this procedure and storing the values returned each iteration. Steps~\ref{step:mh-start} to~\ref{step:mh-end} constitute the collapsed Gibbs move, and steps~\ref{step:gibbs-start} to~\ref{step:gibbs-end} comprise a draw from the constrained parameter conditional.  Each iteration requires a pass of the Kalman filter to calculate the marginal likelihood $\den(\ob{1:\timax}|\lgtm\mcint, \lgtv\mcint)$, plus an additional pass for each Metropolis-Hastings step conducted.

\begin{algorithm}
\small
\onehalfspacing
\begin{algorithmic}[1]
\REQUIRE $\lgtm\mcold, \lgtv\mcold, \ls{1:\timax}\mcold$.
\FOR{each of the Metropolis-Hastings kernels targeting $\postden(\lgtm, \lgtv)$, detailed in sections~\ref{sec:mh-rw-transition-matrix},~\ref{sec:mh-rw-transition-covariance},~and~\ref{sec:rjmcmc-covariance-rank}}
%
\STATE \label{step:mh-start} Propose new parameter values $\lgtm\mcnew, \lgtv\mcnew$.
\STATE With the appropriate acceptance probability, set $\lgtm\mcold \leftarrow \lgtm\mcnew$, $\lgtv\mcold \leftarrow \lgtv\mcnew$.
\STATE \label{step:mh-end} Sample $\ls{1:\timax}\mcold \sim \postden(\ls{1:\timax}|\lgtm\mcold, \lgtv\mcold)$.
%
\STATE \label{step:gibbs-start} Calculate $\tvfull\mcold, \tvrot\mcold$, the Givens decomposition of $\lgtv\mcold$.
\STATE Calculate $\lgtmrot\mcold, \lgtmorth\mcold$, the orthogonal components of $\lgtm\mcold$, using \eqref{eq:transition_matrix_free} \& \eqref{eq:transition_matrix_fixed}.
\STATE Sample $\lgtmrot\mcint, \tvfull\mcint \sim \postden(\lgtmrot,\tvfull|\lgtmorth\mcold,\tvrot\mcold,\ls{1:\timax}\mcold)$, the constrained parameter conditional given in \eqref{eq:constrained_parameter_conditional}.
\STATE \label{step:gibbs-end} Calculate $\lgtm\mcint$ and $\lgtv\mcint$ from $\lgtmrot\mcint, \lgtmorth\mcold, \tvfull\mcint, \tvrot\mcold$ using \eqref{eq:givens_decomposition} \& \eqref{eq:transition_matrix_decomposition}.
%
\ENDFOR
%
\ENSURE $\lgtm\mcint, \lgtv\mcint, \ls{1:\timax}\mcint$.
\end{algorithmic}
\caption{\small MCMC kernel for degenerate linear state space models}
\label{alg:complete-degenerate-model-algorithm}
\end{algorithm}

@@

In the previous sections, we have assumed that $\lgom$ and $\lgov$ are fixed and known a priori. In practice, $\lgov$ will often be unknown. In both the following examples we use models which have,
\begin{IEEEeqnarray}{rCl}
 \lgov &=& \xi_{\ob{}} \idmat     ,
\end{IEEEeqnarray}
%
in which $\xi_{\ob{}}$ is unknown. This additional parameter may be learned in combination with the transition model by the MCMC sampler. This entails changing the target distribution to $\postden(\lgtm, \lgtv, \xi_{\ob{}}, \ls{1:\timax})$. An additional Gibbs move is introduced to draw from $\postden(\xi_{\ob{}} | \lgtm, \lgtv, \ls{1:\timax})$. This may be done exactly if we use a conjugate inverse gamma prior.

@@

The forward-filtering-backward-sampling procedure draws a sample from $\den(\ls{1:\timax}|\ob{1:\timax}, \lgtm, \lgtv)$. This is achieved by first sampling from the final filtering distribution $\den(\ls{\timax}|\ob{1:\timax})$ and then for $\ti=\timax-1 , \dots , 1$ sampling from $\den(\ls{\ti}|\ls{\ti+1}, \ob{1:\ti}, \lgtm, \lgtv)$. Each of these factors is also a Gaussian,
%
\begin{IEEEeqnarray}{l}
 \den(\ls{\ti}|\ls{\ti+1}, \ob{1:\ti}, \lgtm, \lgtv) \nonumber \\
 \quad \propto \den(\ls{\ti+1}|\ls{\ti}, \lgtm, \lgtv) \den(\ls{\ti}| \ob{1:\ti}, \lgtm, \lgtv) \nonumber \\
 \quad = \normalden{\ls{\ti}}{ \lsfmn{\ti} + K_t (\ls{\ti+1}-\lgtm\lsfmn{\ti}) }{ \lspvr{\ti} - K_t \lgtm \lsfvr{\ti} }      ,
\end{IEEEeqnarray}
%
where $K_t = \lsfvr{\ti} \lgtm\tr (\lgtm \lsfvr{\ti} \lgtm\tr + \lgtv)\inv$. The proof is straightforward using standard identities for the Gaussian distribution \cite{Sarkka2013}.
